Advances in Artificial Intelligence for Data Visualization: Automated Reading of Residual Plots with Computer Vision

### 2023-04-19 Milestone_presentation_2/Milestone_presentation_2/Milestone_presentation_2.Rmd
finish slides

@@

@@

# 🔬Visual Inference🔬

@@

- Can you find the **most different plot**?

- Is it the residual plot presented at the previous page?

- If it is not, then the visual pattern is **over-interpreted**!

@@

@@

# 🔬Visual Inference🔬

This procedure is called **visual inference**. It was introduced by Buja, et al. (2009) as an inferential framework to extend confirmatory statistics to visual discoveries.

A **lineup** consists of $m$ randomly placed plots, where one plot is the **data plot** and the remaining $m − 1$ plots (**null plots**) contain data consistent with the null hypothesis.

To perform a **visual test**, observer(s) will be asked to select the **most different plot(s)** from the lineup.

@@

Data used in the $m − 1$ null plots needs to be simulated.

For classical normal linear regression model, the **residual rotation** technique (Buja, et al. 2009) can be applied:

@@

Under the null hypothesis, it is expected that the data plot would have **no distinguishable difference** with the null plots.

@@

Suppose observers are allowed to **select only one plot**. The probability of the observer correctly picks the actual data plot is $1/m$.

@@

However, if we involve $K$ **independent observers** in a visual test, and show them the **same lineup**, there will be possible **dependencies** in the visual test due to **repeated evaluations** of the same lineup.

```{r}
alpha_sample <- rgamma(10, 1, 1)
data.frame(x = 0, y = 0, k = 1:20, alpha = alpha_sample/sum(alpha_sample)) %>%
ggplot() +
  geom_text(aes(x, y, label = paste0("theta[", k, "]~'='~", round(alpha, 2))), parse = TRUE, size = 5) +
  facet_wrap(~k) +
  theme_light(base_size = 12) +
  theme(panel.grid = element_blank(),
        axis.line = element_blank(),
        axis.title = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank())
```

.center[
.caption[
Fig. 5: _Example of the probability of each plot being selected \\(\theta_i\\) under the null hypothesis with parameter \\(\alpha = 1\\)._
]
]

# 📈Statistical Significance📈

VanderPlas et al. (2021) addresses this by modelling the probability of a plot $i$ being selected from a lineup as $\theta_i$:

@@

For **small** $\alpha$:
- Only **a few plots** are attractive and tend to be selected.

For **large** $\alpha$:
- The distribution of the probability of each plot being selected is more **even**.

```{r}
data.frame(k = 1:20, 
           ci = c(1, 0, 0.25, 0, 0, 0.2, 0.45, 0, 5.95, 0, 0, 0, 0, 1.7, 0, 0.2, 0, 1, 0, 0.25)) %>%
ggplot() +
  geom_text(aes(0, 0, label = paste0("c[", k, "]~'='~", ci)), parse = TRUE, size = 5) +
  facet_wrap(~k) +
  theme_light(base_size = 12) +
  theme(panel.grid = element_blank(),
        axis.line = element_blank(),
        axis.title = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank())
```

.center[
.caption[
Fig. 5: _Example of the number of times a plot being selected \\(c_i\\). The sum of all \\(c_i\\) is 11 because the lineup is evaluated by 11 subjects._
]
]

# 📈Statistical Significance📈

The **number of times** plot $i$ being selected in $K$ evaluations is denoted as $c_i$.

@@

- We assume the **individual skill has negligible effect** on the variation of the power to **simplify the model structure**, thereby **obviate a costly large-scale experiment** to estimate complex covariance matrices.

- The **effect size $E$** is the only predictor.
- The intercept is fixed to $log(0.05/0.95)$ so that $\hat{Pr}(\text{reject}~H_0|H_1,E = 0) = 0.05$.

@@

.center[
.caption[
Fig. 5: _Non-linearity forms generated for the residual plots are controlled by the parameter \\(j\\). And the strength of the signal of the non-linearity model is controlled by \\(\sigma\\)._
]
]

@@

```{r out.width = "70%", fig.height = 2.67, fig.width = 8}
set.seed(10086)

@@

.center[
.caption[
Fig. 5: _Heteroskedasticity forms generated for the residual plots are controlled by the parameter \\(a\\). And the strength of the signal of the heteroskedasticity model is controlled by \\(b\\)._
]
]

@@

```{r out.width = "70%"}
magick::image_read_pdf("images/different-n-1.pdf", pages = 1)
```

@@

.center[
.caption[
Fig. 5: _Four different distribution of fitted values that might affect perception of residual plots are used. Three differnet values of \\(n\\) are used to control the signal strength._
]
]

@@

@@

![](images/lineup1.png)

.center[
.caption[
Fig. 5: _Screenshot of the study website._
]
]

@@

@@

@@

.center[
.caption[
Fig. 5: _Comparison of power between different tests for non-linear patterns (uniform fitted values only). The horizontal lines of dots represent non-reject and reject results from visual tests for each lineup. The visual test has multiple power curves estimated from bootstrap samples. The row of scatterplots at the bottom are examples of residual plots corresponding to the specific effect sizes marked by vertical lines in the main plot._
]
]

@@

@@

.center[
.caption[
Fig. 5: _Comparison of power between different tests for heteroskedasticity patterns (uniform fitted values only). The horizontal lines of dots represent non-reject and reject results from visual tests for each lineup. The visual test has multiple power curves estimated from bootstrap samples. The row of scatterplots at the bottom are examples of residual plots corresponding to the specific effect sizes marked by vertical lines in the main plot._
]
]

@@

@@

```{r out.width = "70%"}
magick::image_read_pdf("images/p-value-comparison-1.pdf", pages = 1)
```

@@

.center[
.caption[
Fig. 5: _Rejection rate (\\(p\\)-value \\(\leq0.05\\)) of visual test conditional on the conventional test decision on non-linearity (left) and heteroskedasticity (right) lineups (uniform fitted values only) displayed using a mosaic plot. The visual test rejects less frequently than the conventional test, and (almost) only rejects when the conventional test does. Surprisingly, one lineup in the heteroskedasticity group is rejected by the visual test but NOT the conventional test._
]
]

@@

```{r}
set.seed(10086)

# Base model
mod <- poly_model(sigma = 8)
dat <- mod$gen(1000)
this_p_value <- mod$test(dat, test = "RESET", power = 2:4)
dat %>%
mod$plot(remove_grid_line = TRUE, theme = theme_light(base_size = 12)) +
  theme(axis.text = element_blank()) +
  xlab("Fitted values") +
  ylab("Residuals")
```

@@

.center[
.caption[
Fig. 2: _One classically good looking residual plot but rejected by the RESET test with $p\text{-value} = `r format(round(this_p_value$p_value, digits = 3), nsmall = 3)`$. The model is misspecified but the departure from the model assumption is small._
]
]

@@

With **large sample sizes**, hypothesis tests may reject a residual plot when there is only a **slight departure** (Belsley, Kuh, and Welsch 1980).

<!-- - While such rejections may be **statistically correct**, their **sensitivity** may render the results **impractical**. **Minor defects** in the model are **unlikely** to have a **material impact** and may be best disregarded for practical purposes.  -->

Residual plot diagnostics:
  - **Identify** potential issues that could lead to **incorrect conclusions or errors in subsequent analyses**.

@@

@@

.center[
.caption[
Fig. 5: _The effect of the order of the polynomial on the power of conventional and visual tests. Deeper colour indicates higher order. The default RESET tests under-performs significantly in detecting the "triple-U" shape. To achieve a similar power as other shapes, a higher order polynomial parameter needs to be used for the RESET test, but this higher than the recommended value._
]
]

@@

@@

.center[
.caption[
Fig. 5: _The effect of heteroskedasticity shape (parameter \\(a\\)) on the power of conventional and visual tests. The butterfly has higher power in both tests. Curiously, the visual test has slightly higher power for the "left-triangle" than the "right-triangle" shape, when it would be expected that they should be identical, which is observed in conventional testing._
]
]

@@

class: middle, center

@@

@@

```{r out.width = "60%"}
knitr::include_graphics("images/different-x-dist-poly-power-1.png")
```

@@

.center[
.caption[
Fig. 5: _Comparison of power on lineups with different fitted value distributions for conventional and visual tests for non-linearity and heteroskedasticity patterns. The power curves of conventional tests for non-linearity and heteroskedasticity patterns are produced by RESET tests and BP tests, respectively. Power curves of visual tests are estimated using five evaluations on each lineup. For lineups with a uniform fitted value distribution, the five evaluations are repeatedly sampled from the total eleven evaluations to give multiple power curves (grey). Surprisingly, the fitted value distribution has produces more variability in the power of conventional tests than visual tests. Uneven distributions, normal and lognormal distributions, tend to yield lower power._
]
]

@@

>>>>

# 🔍Motivation and Research Questions🔍

#### Regression experts **consistently recommend plotting residuals for regression diagnostics**, despite the existence of numerous hypothesis test procedures.

#### We would like to **provide evidence** for why this is good advice, using data from a **visual inference experiment**.

#### We would also want to show how **visual inference** can be used to yield reliable and consistent reading of residual plots for better regression diagnostics.

@@

.center[
# 🔬Visual Inference🔬
]

@@

@@

```{r}
set.seed(10131)
mod$gen_lineup(300, k = 20, pos = 8) %>%
  filter(null != FALSE) %>%
  bind_rows(ori_dat %>% mutate(k = 8, null = FALSE)) %>%
  VI_MODEL$plot_lineup(theme = theme_light(base_size = 12),
                       remove_grid_line = TRUE,
                       remove_axis = TRUE)
```

.center[
# 🔬Visual Inference🔬
]

- Please pick the **most different plot**.

```{r}
set.seed(10131)
mod$gen_lineup(300, k = 20, pos = 8) %>%
  filter(null != FALSE) %>%
  bind_rows(ori_dat %>% mutate(k = 8, null = FALSE)) %>%
  VI_MODEL$plot_lineup(theme = theme_light(base_size = 12),
                       remove_grid_line = TRUE,
                       remove_axis = TRUE)
```

.center[
# 🔬Visual Inference🔬
]

- Please pick the **most different plot**.

- Is it plot No.8?

```{r}
set.seed(10131)
mod$gen_lineup(300, k = 20, pos = 8) %>%
  filter(null != FALSE) %>%
  bind_rows(ori_dat %>% mutate(k = 8, null = FALSE)) %>%
  VI_MODEL$plot_lineup(theme = theme_light(base_size = 12),
                       remove_grid_line = TRUE,
                       remove_axis = TRUE)
```

.center[
# 🔬Visual Inference🔬
]

- Please pick the **most different plot**.

- Is it plot No.8?

- If it is not, then the visual pattern is **over-interpreted**! You can not distinguish it from other plots!

```{r}
set.seed(10131)
mod$gen_lineup(300, k = 20, pos = 8) %>%
  filter(null != FALSE) %>%
  bind_rows(ori_dat %>% mutate(k = 8, null = FALSE)) %>%
  VI_MODEL$plot_lineup(theme = theme_light(base_size = 12),
                       remove_grid_line = TRUE,
                       remove_axis = TRUE)
```

.center[
# 🔬Visual Inference🔬
]

- Please pick the **most different plot**.

- Is it plot No.8?

- If it is not, then the visual pattern is **over-interpreted**! You can not distinguish it from other plots!

@@

```{r}
set.seed(10131)
mod$gen_lineup(300, k = 20, pos = 8) %>%
  filter(null != FALSE) %>%
  bind_rows(ori_dat %>% mutate(k = 8, null = FALSE)) %>%
  VI_MODEL$plot_lineup(theme = theme_light(base_size = 12),
                       remove_grid_line = TRUE,
                       remove_axis = TRUE)
```

@@

This framework is called **visual inference**. It was introduced by Buja, et al. (2009) as an inferential framework to extend confirmatory statistics to visual discoveries.

A **lineup** consists of $m$ randomly placed plots, where one plot is the **data plot** and the remaining $m − 1$ plots (**null plots**) containing data consistent with the null hypothesis.

To perform a **visual test**, observer(s) will be asked to select the **most different plot(s)** from the lineup. And we check if the data plot is identified.

@@

Data used in the $m − 1$ null plots needs to be simulated. For **classical normal linear regression model**, the **residual rotation technique** (Buja, et al. 2009) can be applied:

@@

@@

If we involve $K$ **independent observers** in a visual test, and show them the **same lineup**, there will be possible **dependencies** in the visual test due to **repeated evaluations** of the same lineup.

@@

VanderPlas et al. (2021) addresses this by modelling the **probability of a plot $i$ being selected** from a lineup as $\theta_i$:

@@

The **number of times plot $i$ being selected** in $K$ evaluations is denoted as $c_i$.

@@

@@

@@

```{r out.width = "80%", fig.height = 2.67, fig.width = 8}
set.seed(10086)

@@

@@

```{r out.width = "80%"}
magick::image_read_pdf("images/different-n-1.pdf", pages = 1)
```

@@

@@

- We assume the **individual skill has negligible effect** to **simplify the model structure**, thereby **avoid a costly large-scale experiment** to estimate complex covariance matrices.

- The **effect size $E$** is the only predictor.

- The intercept is fixed to $log(0.05/0.95)$ so that $\hat{Pr}(\text{reject}~H_0|H_1,E = 0) = 0.05$.

@@

.center[
# 🌐Study Website🌐
]

```{r out.width = "90%"}
knitr::include_graphics("images/lineup1.png")
```

@@

### (uniform fitted values only)

@@

### Non-linearity Patterns

@@

@@

### Heteroskedasticity Patterns

@@

.center[
# ⚖️Results - Power Comparison of the Tests⚖️
]

The data plot (No.1) is **undistinguishable** from other plots with an extremely small effect size (\\(log_e(E) = -0.48\\)).

The non-linearity pattern is **totally undetectable**.

However, the BP test rejects the pattern with a very small $p\text{-value} = 0.004$. In contrast, the $p\text{-value}$ produced by the visual test is $0.813$.

```{r}
knitr::include_graphics("images/230.png")
```

.center[
# ⚖️Results - Power Comparison of the Tests⚖️
]

Conventional tests are **more sensitive** than the visual test. This is **not necessarily a good feature** for the purposes of diagnosing model defects.

Visual tests:

- Do not require specifying the pattern **ahead of time**.

- Rely purely on **whether the data plot is distinguishable** from "good" residual plots.

- Perform **equally well** regardless of the type of residual departures.

- **Remove any subjective arguments** about whether a pattern is visible or not.

The lineup protocol provides the **calibration for detecting patterns**: that if the pattern in the data plot **can not be distinguished** from patterns in good residual plots, then **no discernible problem** with the model exists.

@@

The visual test rejects **less frequently** than the conventional test, and (almost) **only rejects when the conventional test does**.

@@

```{r out.width = "80%"}
magick::image_read_pdf("images/p-value-comparison-1.pdf", pages = 1)
```

@@

@@

```{r}
knitr::include_graphics("images/331.png")
```

@@

@@

The data plot (No.17) displays a relatively strong heteroskedasticity pattern, and has a strong effect size (\\(log_e(E)=4.02\\)).

This is reflected by the visual test $p\text{-value} = 0.026$.

But the BP test $p\text{-value} = 0.056$, is slightly above the significance cutoff of $0.05$.

Possible reasons:
- Small sample size
- Presence of a few outliers
- Others

@@

The default RESET tests **under-performs significantly** in detecting the "triple-U" shape.

To achieve a similar power as other shapes, a **higher order polynomial parameter** needs to be used for the RESET test, but this higher than the recommended value (4).

@@

@@

The butterfly shape has higher power in both tests.

Curiously, the visual test has **slightly higher power** for the "left-triangle" than the "right-triangle" shape.

@@

@@

@@

.pull-left-center-60[

Surprisingly, the fitted value distribution has produces **more variability** in the power of conventional tests than visual tests.

**Uneven distributions** (normal and lognormal distributions) tend to yield lower power.

.pull-right-140[

@@

```{r}
knitr::include_graphics("images/different-x-dist-poly-power-1.png")
```

@@

@@

1. Conventional residual-based statistical tests are **more sensitive to weak departures** from model assumptions than visual tests as would be evaluated by humans.

2. Conventional tests often reject when departures in the form of non-linearity and heteroskedasticity **are not visibly different** from null residual plots.

3. While conventional tests are correctly detecting the **small but real effects**, human evaluations provide a more **practical** result.

4. Residual plots need to be **delivered as a lineup** to enables a **careful calibration**.

5. Power of the conventional test are **quite sensitive** to **how the predictors are sampled**.

6. The **direction of heteroskedasticity** appears to affect the ability to visually detect it.

---
.center[

# ⏳Future Works⏳

.pull-left[
## Automatic Visual Statistical Inference System

Develop an automatic visual inference system for evaluating lineups of residual plots.

Computer vision models will be trained on simulated data from linear models that violate classical assumptions such as linearity and homoscedasticity.

The performance of the system will be evaluated using the data collected for the first paper.

.pull-right[
## OAVIS: An online automatic visual inference system

Deploy the automatic visual inference as an online application and publish relevant open-source software packages developed during the course of the research. 
]
